Relationships in the extremes and their influence on competition and coexistence

### 2022-07-10 paper/Paper.Rnw
making storage effects singular, as suggested by LMH

%Currently 150 of an allowed 150 words for Ecology Letters
\begin{abstract}
\noindent Species coexistence attracts wide 
interest in ecology. Modern coexistence
theory (MCT) identifies coexistence mechanisms,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures. However, such relationships
are typically measured using covariance, which does not account for 
the possibility that environment and competition may be more related to each other
when they are strong than when weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are common
between ecological variables, and are important for 
extinction risk, ecosystem stability, and other phenomena. 
We extend the MCT, decomposing storage effects to show the influence of ATAs.
Analysis of a simple model
and an empirical example using diatoms
illustrate that ATA influences can be comparable in magnitude to other
mechanisms of coexistence, and that ATAs can make the 
difference between species coexistence and competitive exclusion. 
ATA influences are an important new mechanism of coexistence. 
\end{abstract}

\keywords{asymmetric tail association, biodiversity, coexistence, competition, competitive exclusion, modern coexistence theory, niche differentiation, storage effects}

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Narawani_2013}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effects allow competing species to fluctuate %over time while escaping extinction because storage effects provide 
in abundance without any going extinct by providing
a relative benefit to whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is common than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. 
Third, a buffering mechanism %must exist whereby species limit 
which limits the influence of competition %in times 
when the environment is poor. 
Dormant stages such as seed banks are an example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
rare and common species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. We develop each step of our new theory for a general 
two-species model (below), and also specifically for the lottery model of the Introduction 
(SI section \ref{SI-SIsect:TheoryForLottery}).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption %can be circumvented
is not neccesary, %we make the assumption because 
it simplifies the presentation of new ideas.

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
%Though some readers may find it slightly confusing to simultaneously use the variable $E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like all operators, $\E(\cdot)$ is rendered upright.

We now specify how to decompose storage effects into contributions due to ATAs and 
due to $EC$ correlation \emph{per se}. 
Storage effect for species $i$ was computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which 
%the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
$E$ and $C$
were rendered independent of each other,
for both rare and common species [see (\ref{eq:storage_effects_gen})]. 
Because %these components 
$E$ and $C$ 
were rendered completely independent in the modified model, 
storage effect quantifies the contribution of the whole $EC$ relationship, in all its aspects, 
to the rare/common species growth rate difference. 
But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between %the components 
$E$ and $C$
altered relative to the original distributions.

Specifically, asymmetries of tail association are removed -- see below for how this is done. 
We then define $\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model 
using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, 
(see eqn. (13) in Ellner (2019),
whereas ATAs are %defined in a covariation context
strictly about covariation. 
Therefore, storage effects are the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- storage effects. 
Only storage effects involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is storage effects. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence. To pursue goal G2, below, we will quantify
the magnitude of the new ATA term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms.

@@

%G2 
Fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms), Fig.
\ref{fig:fig3} shows that the ATA term of (\ref{eq:full_decomp}) often, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MCT decomposition. 
Magnitudes of $\Delta_i^{[EC]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels a-d), for which the two species were equal 
competitors, storage effects (blue plus red
lines on Fig. \ref{fig:fig3}) were the only means by which coexistence could occur, and ATA effects
tended to be negative, weakening total storage effects.

@@

%Summary-type paragraph of main results
We produced and applied a new extension of MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables. Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association,
such as ATAs, which we now know are important.

@@

%Differences with the original, analytic version of the MCT
%
%***DAN: I am not so comfortable with this because I have insufficient 
%knowledge of the original MCT. I also do not want to devote the time
%to learning it right now because of other commitments. I've reworded, 
%trying not to change the meaning. Jasmin, how confident are you in 
%these statements? We will also want to try to get Steve Ellner to check 
%them.
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to extend the mathematics of the original MCT to consider ATAs.

>>>>

%Currently 150 of an allowed 150 words for Ecology Letters
\begin{abstract}
\noindent Species coexistence attracts wide 
interest in ecology. Modern coexistence
theory (MCT) identifies coexistence mechanisms,
one of which, the storage effect, hinges on relationships between fluctuations
in environmental and competitive pressures. However, such relationships
are typically measured using covariance, which does not account for 
the possibility that environment and competition may be more related to each other
when they are strong than when weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are common
between ecological variables, and are important for 
extinction risk, ecosystem stability, and other phenomena. 
We extend the MCT, decomposing the storage effect to show the influence of ATAs.
Analysis of a simple model
and an empirical example using diatoms
illustrate that ATA influences can be comparable in magnitude to other
mechanisms of coexistence, and that ATAs can make the 
difference between species coexistence and competitive exclusion. 
ATA influences are an important new mechanism of coexistence. 
\end{abstract}

\keywords{asymmetric tail association, biodiversity, coexistence, competition, competitive exclusion, modern coexistence theory, niche differentiation, storage effect}

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effect'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Narawani_2013}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: the storage effect. We review, conceptually, 
what the storage effect is \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effect allows competing species to fluctuate %over time while escaping extinction because storage effects provide 
in abundance without any going extinct by providing
a relative benefit to whichever species is currently rare; this is the characteristic that makes the storage effect a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effect can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is common than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. 
Third, a buffering mechanism %must exist whereby species limit 
which limits the influence of competition %in times 
when the environment is poor. 
Dormant stages such as seed banks are an example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, the storage effect can be quantified by measuring to what extent EC covariance contributes to differences between 
rare and common species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effect simply because storage effect hinges on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. We develop each step of our new theory for a general 
two-species model (below), and also specifically for the lottery model of the Introduction 
(SI section \ref{SI-SIsect:TheoryForLottery}).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: the storage effect. 
So we begin by defining the storage effect. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption %can be circumvented
is not neccesary, %we make the assumption because 
it simplifies the presentation of new ideas.

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effect for species $i$ is 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
%Though some readers may find it slightly confusing to simultaneously use the variable $E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like all operators, $\E(\cdot)$ is rendered upright.

We now specify how to decompose storage effect into contributions due to ATAs and 
due to $EC$ correlation \emph{per se}. 
Storage effect for species $i$ was computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which 
%the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
$E$ and $C$
were rendered independent of each other,
for both rare and common species [see (\ref{eq:storage_effects_gen})]. 
Because %these components 
$E$ and $C$ 
were rendered completely independent in the modified model, 
storage effect quantifies the contribution of the whole $EC$ relationship, in all its aspects, 
to the rare/common species growth rate difference. 
But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between %the components 
$E$ and $C$
altered relative to the original distributions.

Specifically, asymmetries of tail association are removed -- see below for how this is done. 
We then define $\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model 
using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effect due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The term $\Delta_i^{[E \nat C]}$ is the component of storage effect due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effect, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, 
(see eqn. (13) in Ellner (2019),
whereas ATAs are %defined in a covariation context
strictly about covariation. 
Therefore, the storage effect is the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is similar to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- the storage effect. 
Only storage effect involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is the storage effect. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. One typically (but see below) must examine this quantity for each species, $i$,
to obtain complete information about coexistence. To pursue goal G2, below, we will quantify
the magnitude of the new ATA term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms.

@@

%G2 
Fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms), Fig.
\ref{fig:fig3} shows that the ATA term of (\ref{eq:full_decomp}) often, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MCT decomposition. 
Magnitudes of $\Delta_i^{[EC]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels a-d), for which the two species were equal 
competitors, storage effect (blue plus red
lines on Fig. \ref{fig:fig3}) was the only means by which coexistence could occur, and ATA effects
tended to be negative, weakening total storage effect.

@@

%Summary-type paragraph of main results
We produced and applied a new extension of MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed the storage effect into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to the storage effect and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables. Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effect is typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effect should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association,
such as ATAs, which we now know are important.

@@

%Differences with the original, analytic version of the MCT
%
%***DAN: I am not so comfortable with this because I have insufficient 
%knowledge of the original MCT. I also do not want to devote the time
%to learning it right now because of other commitments. I've reworded, 
%trying not to change the meaning. Jasmin, how confident are you in 
%these statements? We will also want to try to get Steve Ellner to check 
%them.
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effect as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to extend the mathematics of the original MCT to consider ATAs.
